Generative AI for image reconstruction in Intensity Interferometry: a first attempt

### 2025-11-06 Phase_II/draft/intro_reformatted.tex
intro_reformatted, II.tex, refs.bib, Paragan.tex

\section{Introduction}
The humanity instinctively feels a symbiotic relationship with the stars. One of the primary scientific projects of the humanity is to figure out what the stars are and how do they do what they do. The first obvious step in this project, beyond measuring their global parameters like diameter, mass, orbital and astromteric elements, is to obtain images of the stars with all details of the stellar surfaces as we do routinely in case of the Sun. In other words, the objective is to achieve capability of high fidelity image reconstruction of distant stars. Two interferometry based techniques, namely, Michelson Interferometry (MI) or Intensity Interferometry (II) have emerged during the last century to address this objective. A discussion of the development of these two approaches and comparison of their respective merits and challenges can be found in \citep{Rai2025}. The work reported here presents the results of a first effort at applying a conditional Generative Adversarial (neural) Network (cGAN) to image reconstruction of a fast rotator using its simulated II observations.

@@

We report here, the results of our attempt, the first of its kind, to reconstruct the II-simulated gravity-darkened images of fast rotating stars using a cGAN neural network architecture. Image reconstruction in gravity darkened fast rotating stars has long been examined using various methods in MI \citep{vanBelle2001, DomicianodeSouza2003, DomicianodeSouza2005, Monnier2007, Pedretti2009, Zhao2009, Martinez2021}. Recently photosphere oblateness of $\gamma$-Cassiopeia \citep{2025arXiv250615027A} has been measured at the VERITAS observatory using II. These results, and especially, that by Archer et al. \cite{2025arXiv250615027A}  put our work in context. Thus our simulation is a natural next step of the work of Archer et al and others. We emulate the cGAN model of Isola et al. \cite{isola2017image} to reconstruct images of fast-rotating stars using their simulated Intensity Interferograms and simulated sky-intensity distributions as input data for training, testing, and validation. We consider four Imaging Cherenkov Telescope Arrays (IACTs) and simulate observation of a fast-rotating star. The image predicted by the trained GAN shows promising results in reconstructing the star’s shape and size. The reconstructed brightness distributions are then assessed using moments.

This paper is organized as follows. The next section discusses briefly the past efforts at image reconstruction on II, followed by the section on a discussion of Intensity Interferometry, focusing on its signal and noise characteristics for fast-rotating stars along the Earth’s rotation. The following section introduces the GAN formulation and its structure. The fifth section details the parameter selection for training the GAN for image reconstruction. The sixth section presents the results of the trained GAN both visually and via image moments. Finally, the paper concludes with a discussion of the overall results.
\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{fig/telescope.png}
	\caption{The telescope configuration with similar properties each, where diameter is 12 meter and baselines are spreaded over more than 100 meter. This setup is used to simulate the signal for II observation.}
	\label{fig:teles}
\end{figure}

>>>>

\section{Introduction}
The humanity instinctively feels a symbiotic relationship with the stars. One of the primary scientific projects of the humanity is to figure out what the stars are and how do they do what they do. The first obvious step in this project, beyond measuring their global parameters like diameter, mass, orbital and astrometric elements, is to obtain images of the stars with all details of the stellar surfaces. In case of the Sun, this is done routinely by observatories and sun-observation satellites. But such routine still remains a challenge even for the $\alpha$-centauri system, our nearest stellar neighbour. The objective is to achieve capability of high fidelity image reconstruction of distant stars. Two interferometry based techniques, namely, Michelson Interferometry (MI) or Intensity Interferometry (II) have emerged during the last century to address this objective. A discussion of the development of these two approaches and comparison of their respective merits and challenges can be found in \citep{Rai2025}. The work reported here presents the results of a first effort at applying a Conditional Generative Adversarial (neural) Network (C-GAN) to image reconstruction of a fast rotator using its simulated II observations.

@@

We report here, the results of our attempt, the first of its kind, to reconstruct the II-simulated gravity-darkened images of fast rotating stars using a C-GAN neural network architecture. Image reconstruction in gravity darkened fast rotating stars has long been examined using various methods in MI \citep{vanBelle2001, DomicianodeSouza2003, DomicianodeSouza2005, mcalister2005first, Monnier2007, Pedretti2009, Zhao2009, Martinez2021}. Recently photosphere oblateness of $\gamma$-Cassiopeia \citep{Archer-arXiv-2025} has been measured at the VERITAS observatory using II. These results, and especially, that by \cite{Archer-arXiv-2025} put our work in context. Thus our simulation is a natural next step of the work of Archer et al and others. We emulate the C-GAN model of Isola et al. \cite{isola2017image} to reconstruct images of fast-rotating stars using their simulated Intensity Interferograms and simulated sky-intensity distributions as input data for training, testing, and validation. We consider an array of four Imaging Cherenkov Telescopes (IACTs) and simulate observation of a fast-rotating star. The image predicted by the trained C-GAN shows promising results in reconstructing the star’s shape and size. The reconstructed brightness distributions are then assessed using moments.

This paper is organized as follows. The next section discusses briefly the past efforts at image reconstruction on II, followed by the section on a discussion of Intensity Interferometry, focusing on its signal and noise characteristics for fast-rotating stars along the Earth’s rotation. The following section introduces the C-GAN formulation and its structure. The fifth section details the parameter selection for training the network for image reconstruction. The sixth section presents the results of the trained network both visually and via image moments. Finally, the paper concludes with a discussion of the overall results.
\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{fig/telescope.png}
	\caption{A schematic representation of {\textcolor{blue}{a hypothetical observation facility with an the array of four Cherenkov Telescopes (IACTs). This array is used in simulating the Intensity Interferometry (II) observation of the fast rotators.}} Each of the telescopes has a diameter of 12m. The baselines provided by the array are of the order of 100m.}
	\label{fig:teles}
\end{figure}
